Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriterial problems

### 2024-07-05 2024/OPTIMA/OPTIMA_2024_MCO.tex
OPTIMA fix bug
fix some mistakes

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%\urlstyle{rm}
%
\begin{document}
%
\title{Comparative efficiency of machine learning models for enhancing algorithms in solving multiextremal multicriteria problems\thanks{This work was supported by the Ministry of Science and Higher Education of the Russian Federation, project no. FSWR-2023-0034, and by by the Research and Education Mathematics Centre ``Mathematics for Future Technologies'', project no. 075-02-2024-1439.} }
%
\titlerunning{Algorithms for solving multicriteria problems}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Sergey Konnov\orcidID{0009-0003-4590-0870} \and
Evgeny Kozinov \Letter\orcidID{0000-0001-6776-0096} \and
Konstantin Barkalov \Letter\orcidID{0000-0001-5273-2471} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670}}
%
\authorrunning{S. Konnov, E. Kozinov, K. Barkalov and V. Grishagin}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.

@@

\begin{equation}
\label{eq:01}
  \Phi(y) = (f_1 (y),f_2 (y), \dots, f_s(y))
\end{equation}
where scalar functions $f_i(y)$, $1 \leq i \leq s$, are considered as the partial efficiency criteria, $y=(y_1,y_2, \dots ,y_N)$ is the vector of independent parameters belonging to the feasible domain $D \in R^N$ that is the hyperparallelepiped
\begin{equation}
\label{eq:02}
    D=\{y \in R^N : a_i \leq y_i \leq b_i, 1 \leq i \leq N\}
\end{equation}
in the $N$-dimensional Euclidean space, and $a_i$, $b_i$, $1 \leq i \leq N$, are constant values.
The best value of the partial criterion $f_i(y)$, $1 \leq i \leq s$, is supposed to correspond to the global minimum of this criterion in the domain $D$.

We will denote this problem in the form
\begin{equation}
\label{eq:02a}
  \Phi(y) = (f_1 (y),f_2 (y), \dots, f_s(y)) \to \min, y \in D
\end{equation}

@@

\begin{equation}
\label{eq:03}
|f_i (y') - f_i (y'')| \leq L_i \|y' - y''\| ,y',y'' \in D, 1 \leq i \leq s,
\end{equation}
where Lipschitz constants $L_i>0$, $1 \leq i \leq s$, are unknown \textit{a priopi}.

@@

For solving the optimization problem (\ref{eq:02a}), in our study the computational scheme \cite{ML_MCO_2023} has been applied. It consists of the following main stages.
 \begin{enumerate}
	\item Scalarization of the vector criterion (\ref{eq:01}).
	\item Dimensionality reduction.
	\item Optimization of transformed functions.
\end{enumerate}
Below is a brief description of each stage.

@@

It should be noted that if the convolution $F(\lambda,y)$ satisfies the Lipschitz condition then the reduced function $\varphi(y(x)) = F(\lambda,y(x))$ meets the H{\" o}lder inequality
\begin{equation}
\label{eq:08}
|\varphi(y(x')) - \varphi(y(x''))| \leq H \|x' - x''\|^{1/N} , x \in [0,1],
\end{equation}
with the H{\" o}lder constant $H=2L\sqrt{N+3}$, where $L$ is the Lipschitz constant of $F(\lambda,y)$ in the domain $D$.

@@

\textit{Rule 1.} Renumber by subscripts all trial points in increasing order
\begin{equation}
    \label{eq:09}
    0 = x_0 < x_1 < \dots < x_i < \dots < x_{k-1} = 1.
\end{equation}
and juxtapose to them the values $z_i=\varphi(y(x_i))$, $0 \leq i \leq k-1$.

@@

\begin{equation}
    \label{eq:11}
    R(i) = \varrho_i + \frac{(z_i-z_{i-1})^2}{m^2 \varrho_i} - \frac{2 (z_i+z_{i-1})}{m}, 1 \leq i \leq k-1.
\end{equation}
called \textit{characteristic} of the subinterval $(x_{i-1} ,x_i)$, $1 \leq i \leq k-1$.

\textit{Rule 4.} Choose the subinterval $(x_{t-1} ,x_t)$ with maximal characteristic $R(t)$
\begin{equation}
    \label{eq:12}
    R(t) = \max_{1 \leq i \leq k-1} {R(i)}.
\end{equation}

\textit{Rule 5.} Accept
\begin{equation}
    \label{eq:13}
    x^{k} = \frac{x_t + x_{t-1}}{2} - sign(z_t - z_{t-1}) \frac{1}{2r} \left(\frac{|z_t - z_{t-1}|}{m} \right)^N,
\end{equation}
as the point for the next trial.

The process of finding a solution to the current problem (\ref{eq:07}) is terminated when the condition
\begin{equation}
    \label{eq:14}
    \varrho < \varepsilon.
\end{equation}
is met where $\varepsilon > 0$ is a preassigned search accuracy.

@@

During minimization of one problem (\ref{eq:07}) with corresponding weight vector $\lambda$, the trial points $x^0, x^1,\dots,x^{k-1}$ from the interval $[0,1]$ generate points $y^0, y^1,\dots,y^{k-1}$ in the domain $D$, where $y^i=y(x^i)$, $0 \leq i \leq k-1$, and $y(x)$ is the Peano mapping. After computing values $f^i_j=f_j(y^i)$, $0 \leq i \leq k-1$, $1 \leq j \leq s$, of the partial criteria at points $y^i$ one can organize the set 
\begin{equation}
    \label{eq:15}
    \Omega_K=\{(x^q,y^q,f^q_1,\dots,f^q_s)^T: 0 \leq q \leq K\}.
\end{equation}
called \textit{Search State Matrix} (SSM).

@@

The second approach also aims to further decreasing the number of trials required to find the Pareto set. It suggests to change the form of the characteristics (\ref{eq:11}) in the GSA scheme. It is proposed to use the characteristic 
\begin{equation}
    \label{eq:16}
    R(i) = R_{gsa} (i) +  \alpha R_{ps} (i).
\end{equation}
instead of the characteristic (\ref{eq:11}) where $R_{gsa}(i)$ coincides with the expression (\ref{eq:11}) and the additional term $R_{ps}(i)$ depends on closeness of the points $y(x_{i-1})$, $y(x_{i})$ to the Pareto set, namely, the closer these points to the Pareto set, the greater the value $R_{ps}(i)$.

@@

Fig. \ref{fig:1} shows the behavior of the characteristic $R_{ps}(i)$ in the criteria space for different models. Red points are related to the Pareto set, orange those to the class  where points are not effective ones.

@@

Computations were conducted on the ``Lobachevsky'' supercomputer node (two Intel Sandy Bridge E5-2660 2.2 GHz processors and 64 Gb RAM) using the scikit-learn 0.24.2 library.
The proposed algorithms were implemented in Python in intelligent optimization framework iOpt \cite{iOpt_url}.

>>>>

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%\urlstyle{rm}
%
\begin{document}
%
\title{Comparative efficiency of machine learning models for enhancing algorithms in solving multiextremal multicriteria problems\thanks{This work was supported by the Ministry of Science and Higher Education of the Russian Federation, project no. FSWR-2023-0034, and by by the Research and Education Mathematics Centre ``Mathematics for Future Technologies'', project no. 075-02-2024-1439.} }
%
\titlerunning{Algorithms for solving multicriteria problems}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Sergey Konnov\orcidID{0009-0003-4590-0870} \and
Evgeny Kozinov \Letter\orcidID{0000-0001-6776-0096} \and
Konstantin Barkalov \orcidID{0000-0001-5273-2471} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670}}
%
\authorrunning{S. Konnov, E. Kozinov, K. Barkalov and V. Grishagin}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.

@@

\begin{equation}
\label{eq:01}
  \Phi(y) = (f_1 (y),f_2 (y), \dots, f_s(y))
\end{equation}
where scalar functions $f_i(y)$, $1 \leq i \leq s$, are considered as the partial efficiency criteria, $y=(y_1,y_2, \dots ,y_N)$ is the vector of independent parameters belonging to the feasible domain $D \in R^N$ that is the hyperparallelepiped
\begin{equation}
\label{eq:02}
    D=\{y \in R^N : a_i \leq y_i \leq b_i, 1 \leq i \leq N\}
\end{equation}
in the $N$-dimensional Euclidean space, and $a_i$, $b_i$, $1 \leq i \leq N$, are constant values.
The best value of the partial criterion $f_i(y)$, $1 \leq i \leq s$, is supposed to correspond to the global minimum of this criterion in the domain $D$.
Let's write this problem in the form
\begin{equation}
\label{eq:02a}
  \Phi(y) = (f_1 (y),f_2 (y), \dots, f_s(y)) \to \min, \; y \in D.
\end{equation}

@@

\begin{equation}
\label{eq:03}
|f_i (y') - f_i (y'')| \leq L_i \|y' - y''\| ,y',y'' \in D, 1 \leq i \leq s,
\end{equation}
where Lipschitz constants $L_i>0$, $1 \leq i \leq s$, are unknown a priopi.

@@

For solving the optimization problem (\ref{eq:02a}), in our study the computational scheme from \cite{ML_MCO_2023} has been applied. It consists of the following main stages.
 \begin{enumerate}
	\item Scalarization of the vector criterion (\ref{eq:01}).
	\item Dimensionality reduction.
	\item Optimization of transformed functions.
\end{enumerate}
Below is a brief description of each stage.

@@

It should be noted that if the convolution $F(\lambda,y)$ satisfies the Lipschitz condition then the reduced function $\varphi(y(x)) = F(\lambda,y(x))$ meets the H{\" o}lder inequality
\begin{equation}
\label{eq:08}
|\varphi(y(x')) - \varphi(y(x''))| \leq H |x' - x''|^{1/N} , x \in [0,1],
\end{equation}
with the H{\" o}lder constant $H=2L\sqrt{N+3}$ where $L$ is the Lipschitz constant of $F(\lambda,y)$ in the domain $D$.

@@

\textit{Rule 1.} Renumber by subscripts all trial points in increasing order
\begin{equation}
    \label{eq:09}
    0 = x_0 < x_1 < \dots < x_i < \dots < x_{k-1} = 1
\end{equation}
and juxtapose to them the values $z_i=\varphi(y(x_i))$, $0 \leq i \leq k-1$.

@@

\begin{equation}
    \label{eq:11}
    R(i) = \varrho_i + \frac{(z_i-z_{i-1})^2}{m^2 \varrho_i} - \frac{2 (z_i+z_{i-1})}{m}, 1 \leq i \leq k-1,
\end{equation}
called \textit{characteristic} of the subinterval $(x_{i-1} ,x_i)$, $1 \leq i \leq k-1$.

\textit{Rule 4.} Choose the subinterval $(x_{t-1} ,x_t)$ with maximum characteristic $R(t)$
\begin{equation}
    \label{eq:12}
    R(t) = \max_{1 \leq i \leq k-1} {R(i)}.
\end{equation}

\textit{Rule 5.} Accept
\begin{equation}
    \label{eq:13}
    x^{k} = \frac{x_t + x_{t-1}}{2} - sign(z_t - z_{t-1}) \frac{1}{2r} \left(\frac{|z_t - z_{t-1}|}{m} \right)^N
\end{equation}
as the point for the next trial.

The process of finding a solution to the current problem (\ref{eq:07}) is terminated when the condition
\begin{equation}
    \label{eq:14}
    \varrho_t < \varepsilon
\end{equation}
is met where $\varepsilon > 0$ is a preassigned search accuracy.

@@

During minimization of one problem (\ref{eq:07}) with corresponding weight vector $\lambda$, the trial points $x^0, x^1,\dots,x^{k-1}$ from the interval $[0,1]$ generate points $y^0, y^1,\dots,y^{k-1}$ in the domain $D$, where $y^i=y(x^i)$, $0 \leq i \leq k-1$, and $y(x)$ is the Peano mapping. After computing values $f^i_j=f_j(y^i)$, $0 \leq i \leq k-1$, $1 \leq j \leq s$, of the partial criteria at points $y^i$ one can organize the set 
\begin{equation}
    \label{eq:15}
    \Omega_K=\{(x^q,y^q,f^q_1,\dots,f^q_s)^T: 0 \leq q \leq K\}
\end{equation}
called \textit{Search State Matrix} (SSM).

@@

The second approach also aims to further decreasing the number of trials required to find the Pareto set. It suggests to change the form of the characteristics (\ref{eq:11}) in the GSA scheme. It is proposed to use the characteristic 
\begin{equation}
    \label{eq:16}
    R(i) = R_{gsa} (i) +  \alpha R_{ps} (i)
\end{equation}
instead of the characteristic (\ref{eq:11}), where $R_{gsa}(i)$ coincides with the expression (\ref{eq:11}) and the additional term $R_{ps}(i)$ depends on closeness of the points $y(x_{i-1})$, $y(x_{i})$ to the Pareto set, namely, the closer these points to the Pareto set, the greater the value $R_{ps}(i)$.

@@

Fig. \ref{fig:1} shows the behavior of the characteristic $R_{ps}(i)$ in the criteria space for different models. Red points are related to the Pareto set, orange ones belong to the class of inefficient points.

@@

Computations were conducted on the ``Lobachevsky'' supercomputer node (two Intel Sandy Bridge E5-2660 2.2 GHz processors and 64 Gb RAM) using the scikit-learn 0.24.2 library.
The proposed algorithms were implemented in Python in the \textit{i}ntelligent \textit{opt}imization framework iOpt \cite{iOpt_url}.
